---
title: "Unsupervised Wisdom from Medical Narratives"
subtitle: "Exploring Medical Narratives on Older Adult Falls"
author: "Ifechukwu Mbamali"
date: last-modified
date-format: "DD-MM-YYYY"
description: "Exploring unsupervised machine learning methods to extract insights from medical narratives about older adults (age 65+) fall"
format: 
  html:
    theme: default
    code-fold: true
    embed-resources: true
    toc-location: left
self-contained: true
editor: visual
toc: true
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged

title-block-banner: "#2274cc"
reference-location: margin
---

```{r libraries, message=FALSE, warning=FALSE}
#| echo: false
library(ggplot2)
library(tidymodels)
library(tidyverse)
library(vroom)
library(LSAfun)
library(jsonlite)
library(tidyclust)
library(SnowballC) #NLP Stemming
library(tidytext)
library(visNetwork)
library(igraph)
library(hunspell) #spell checker - NLP stemming
library(arrow) #read parquet.gzip file
library(data.table)
library(plotly)
#devtools::install_github("ricardo-bion/ggradar")
library(ggradar)
library(DT)
library(reactable)
library(formattable)
library(tidygraph)
library(embed)
library(janitor)
library(fmsb)
library(scales)
library(textrank)
library(factoextra)
library(dbscan) #dbscan library
library(purrr)  
library(cluster)
```

# Introduction

The analysis made use of embeddings, dimensionality reduction, clustering algorithms, network graphs and text summarization techniques to effectively identify, and understand themes from medical narratives on older adults falls.

**Key findings:** The use of embeddings in combination with dimensionality reduction techniques proved effective in extracting cluster themes. DBSCAN[^1] outperforms k-means in cluster identification. Patients in the "*Alcohol-Related Head Injuries and Falls*" group tend to be younger, while the "*Atrial fibrillation related falls*" group was generally older, and the "*Syncope-Related Head* *Injuries*" group had a higher rate of severe cases compared to others. In comparison to the previous year (2021), cases involving "*Head Injuries from Falls*", "*Syncope-Related Head* *Injuries*" and "*Rib Injuries from Falls*" saw the most significant increase in the average number of cases.

[^1]: Density-based clustering algorithm, introduced in Ester et al. 1996, which can be used to identify clusters of any shape in data set containing noise and outliers. DBSCAN stands for Density-Based Spatial Clustering and Application with Noise

Ultimately, insights gained through this analysis can help inform policies and interventions to reduce older adult falls. [Competition](https://www.drivendata.org/competitions/217/cdc-fall-narratives/page/763/#submission-format) hosted by Centers for Disease Control and Prevention.

# Data Overview

The analysis made use of 2 data-sets:

-   Primary data-set
-   OpenAI embeddings data-set

::: {.callout-note collapse="true"}
## Loading Data

```{r}
#recode the encoded variables in the dataset to human-readable values
mapping <- fromJSON("data/variable_mapping.json")
# Convert to data frames so we can use in joins
mapping_tables <- list()
for (col in names(mapping)) {
    mapping_tables[[col]] <- data.frame(
        ind=as.integer(names(mapping[[col]])),  # change to integer types
        values=unlist(mapping[[col]])
    )
}

```

```{r}
# Load primary data
pdf <- read.csv("data/primary_data.csv" )

# Join and replace encoded column
for (col in names(mapping)) {
    pdf <- pdf %>%
        left_join(mapping_tables[[col]], by=setNames("ind", col)) %>%
        mutate(!!col := values) %>%
        select(-values)
}
```

```{r}
#| echo: false

#################################
## speed-up notebook rendering ##
#################################


##run once or alternatively load "data/embeddings.csv"

# emb<- read_parquet("openai_embeddings_primary_narratives.parquet.gzip")

## Splitting the "embedding" column into individual columns
#note large file takes a long time 115128 by 1537. 

# emb2 <- emb %>%
# 
#   unnest_wider(embedding, names_sep = "_")
# 
# fwrite(emb2, "data/embeddings.csv")

#load embeddings
emb2b = fread("data/embeddings.csv")
```

```{r}
emb2c = emb2b |> select(1:51)
```

```{r}
#| echo: false


#################################
## speed-up notebook rendering ##
#################################

##run once or alternatively load "data/embeddings_pca.csv"

##Apply dimentionality reduction step


# emb2d = recipe(~.,-cpsc_case_number, data = emb2b)|>
#   step_rm(cpsc_case_number)|>
#   step_normalize(all_numeric_predictors())|>
#   step_pca(all_numeric_predictors()) |> #, threshold = 0.8
#   prep()|>
#   juice()
# fwrite(emb2d, "data/embeddings_pca.csv") 
```

```{r}

#################################
## speed-up notebook rendering ##
#################################

emb2d = fread("data/embeddings_pca.csv")#embeddings_pca_x.csv
```

```{r}
emb2d  = emb2b|>
  select(cpsc_case_number)|>
  bind_cols(emb2d )
```
:::

::: panel-tabset
## Narratives Data

```{r}
as.datatable(formattable(pdf|>
                           head(1)
                         ), rownames = F,
filter = 'top', 
options = list(
  pageLength = 10, autoWidth = F,
  order = list(list(2, 'desc'))#asc
),
class = 'bootstrap'
)
```

## Embeddings Data

```{r}
#table view of the first 51 columns of the raw embeddings file
as.datatable(formattable(emb2c|>
                           head(5)
                         ), rownames = F,
filter = 'top',
options = list(
  pageLength = 10, autoWidth = F,
  order = list(list(2, 'desc'))#asc
),
class = 'bootstrap'
)
```
:::

# Text cleaning and pre-processing

Some of the general preprocessing steps include:

-   **New fields:** Additional columns were introduced to the primary data-set to determine severity levels based on disposition column and another to categorize activities based on the narratives.
-   **Replacing Medical Abbreviations:** The narrative column was also processed by replacing abbreviations with full clinical definition to improve readability.

::: {.callout-note collapse="true"}
-   Creating new columns: Severity Level and Activity

```{r}
#create a column called "severity_level" that says "severe" if the number 4 or 5 is contained in the "disposition" column, and "not severe" otherwise
pdf <- pdf |>
  mutate(severity_level = ifelse(grepl("4|5", disposition), "severe", "not severe"))

#create a column called "activity" that captures the text between "-" and "(" or "-" and "," if the term "ACTIVITY" is contained in the "product_1" column, and "others" otherwise

pdf <- pdf |>
  mutate(activity = ifelse(grepl("ACTIVITY", product_1),
                           sub(".*-(.*?)[(,].*", "\\1", product_1),
                           "others"))

#modifies the "activity" column by replacing "others" with "fainted" if "SYNCOPAL" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("SYNCOPAL|DIZZY|WEAK|WEAKNESS|SYNCOPE", narrative), "fainted", activity))

#modifies the "activity" column by replacing "others" with "WALKING" if "WALKING" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("WALKING|WALK", narrative), "WALKING", activity))

#modifies the "activity" column by replacing "others" with "STANDING" if "STANDING" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("STANDING|STAND", narrative), "STANDING", activity))

#modifies the "activity" column by replacing "others" with "SITTING" if "SITTING" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("SITTING|SIT", narrative), "SITTING", activity))

#modifies the "activity" column by replacing "others" with "Stair Navigation" if "FLIGHT" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("FLIGHT|STAIRS", narrative), "Stair Navigation", activity))

#modifies the "activity" column by replacing "others" with "RISING" if "FLIGHT" is contained in the "narrative" column
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("GETTING|CHAIR|BED|STOOD UP", narrative), "RISING", activity))

#---
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("SLIPPED|SLIP", narrative), "SLIPPED", activity))

#---
pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("TRIPPED|TRIP", narrative), "TRIPPED", activity))

#---

pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("BENDING|BENT|BEND|PICK UP|PICKING UP", narrative), "BENDING", activity))

#---

pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("MECH|MECHANICAL", narrative), "MECHANICAL", activity))

#---

pdf <- pdf |>
  mutate(activity = ifelse(activity == "others" & grepl("LOST BALANCE", narrative), "LOST BALANCE", activity))

#---

pdf <- pdf |>
  mutate(activity = ifelse(grepl("BASKETBALL|BASEBALL|BALL|SPORTS|SPORT|BILLIARDS|BOWLING|SKATING|GOLF|TENNIS|MOUNTAIN CLIMBING|SKIING|SOCCER|HOCKEY|FISHING|SWIMMING|MARTIAL ARTS|LACROSSE|TUBING|HORSEBACK RIDING|SURFING|WRESTLING|BADMINTON|SHUFFLEBOARD|FENCING", activity), "SPORTS", activity))
```

-   Cleaning Narratives by replacing medical & other abbreviations

```{r}
# Define the medical_terms dictionary
medical_terms <- list(
  "&" = "and",
  "***" = "",
  ">>" = "clinical diagnosis",
  "@" = "at",
  "abd" = "abdomen",
  "af" = "accidental fall",
  "afib" = "atrial fibrillation",
  "aki" = "acute kidney injury",
  "am" = "morning",
  "ams" = "altered mental status",
  "bac" = "blood alcohol content",
  "bal" = "blood alcohol level,",
  "biba" = "brought in by ambulance",
  "c/o" = "complains of",
  "chi" = "closed-head injury",
  "clsd" = "closed",
  "cpk" = "creatine phosphokinase",
  "cva" = "cerebral vascular accident",
  "dx" = "diagnosis",
  "ecf" = "extended-care facility",
  "er" = "emergency room",
  "etoh" = "ethyl alcohol",
  "eval" = "evaluation",
  "fib" = "fibrillation",
  "fd" = "fall detected",
  "fx" = "fracture",
  "fxs" = "fractures",
  "glf" = "ground level fall",
  "h/o" = "history of",
  "htn" = "hypertension",
  "hx" = "history of",
  "inj" = "injury",
  "inr" = "international normalized ratio",
  "intox" = "intoxication",
  "l" = "left",
  "lac" = "laceration",
  "loc" = "loss of consciousness",
  "lt" = "left",
  "mech" = "mechanical",
  "mult" = "multiple",
  "n.h." = "nursing home",
  "nh" = "nursing home",
  "p/w" = "presents with",
  "pm" = "afternoon",
  "pt" = "patient",
  "pta" = "prior to arrival",
  "pts" = "patient's",
  "px" = "physical examination",
  "r" = "right",
  "r/o" = "rules out",
  "rt" = "right",
  "s'd&f" = "slipped and fell",
  "s/p" = "after",
  "sah" = "subarachnoid hemorrhage",
  "sdh" = "acute subdural hematoma",
  "sts" = "sit-to-stand",
  "t'd&f" = "tripped and fell",
  "tr" = "trauma",
  "uti" = "urinary tract infection",
  "w/" = "with",
  "w/o" = "without",
  "wks" = "weeks"
)

# Define the clean_narrative function
clean_narrative <- function(text) {
  # Convert text to lowercase
  text <- tolower(text)

  # Define regex pattern for DX
  regex_dx <- "([\\W]*(dx)[\\W]*)"
  text <- gsub(regex_dx, ". dx: ", text)

  # Define regex pattern for age and sex
  regex_age_sex <- "(\\d+)\\s*?(yof|yf|yo\\s*female|yo\\s*f|yom|ym|yo\\s*male|yo\\s*m)"
  age_sex_match <- regexpr(regex_age_sex, text)

  # Format age and sex
  if (age_sex_match > 0) {
    age <- regmatches(text, age_sex_match)[[1]][1]
    sex <- regmatches(text, age_sex_match)[[1]][2]

    if ("f" %in% sex) {
      text <- gsub(age_sex_match, "patient", text)
    } else if ("m" %in% sex) {
      text <- gsub(age_sex_match, "patient", text)
    }
  }

  # Translate medical terms
  for (term in names(medical_terms)) {
    if (term %in% c("@", ">>", "&", "***")) {
      pattern <- paste0("(", gsub("[*]", "[*]", term), ")")
      text <- gsub(pattern, paste0(" ", medical_terms[[term]], " "), text)
    } else {
      pattern <- paste0("\\b(", gsub("[*]", "[*]", term), ")\\b")
      text <- gsub(pattern, medical_terms[[term]], text)
    }
  }

  # Capitalize sentences
  text <- gsub("(^|\\.[[:space:]]+)([a-z])", "\\1\\U\\2", text, perl = TRUE)

  # Convert text to uppercase
  #text <- toupper(text)

  return(text)
}

# Test the function
input_text <- "The pt is a 45 yof who c/o abdominal pain. Dx: uti. She fell and has a left hip fx."
cleaned_text <- clean_narrative(input_text)
cat(cleaned_text)
```

```{r}
##############################
## speed notebook rendering ##
##############################

##run once or alternatively load "data/clean_narrative_data.csv"

## applying cleaning function to data

# pdf$narrative_orig = pdf$narrative
# pdf_0 <- pdf %>%
#   mutate(narrative = map_chr(narrative, clean_narrative))
```

```{r}
##############################
## speed notebook rendering ##
##############################

## speed up render by saving file to excel and loading it up.
#fwrite(pdf_0, "data/clean_narrative_data.csv")
```

```{r}
##############################
## speed notebook rendering ##
##############################
pdf_0 = fread("data/clean_narrative_data.csv")
```
:::

# General Analysis of Narratives

In this section, text processing and text analysis tasks were performed on the cleaned narrative column.The code takes text data, removes certain specified words[^2] and stop words, tokenizes it into bigrams, counts the frequency of these bigrams, and calculates the percentage of occurrence for each bigram while performing various text cleaning and filtering operations along the way.

[^2]: re-occurring words that do not provide any insightful information e.g"yom", "yof" etc

::: {.callout-note collapse="true"}
```{r}
#https://paldhous.github.io/NICAR/2019/r-text-analysis.html
pdf5 = pdf_0 %>%
  #filter(activity=="Stair Navigation")|>
  mutate(narrative = 
           gsub("\\bYOF\\b|\\bYOM\\b|\\bPT\\b|\\bDX\\b|\\byom\\b|\\byof\\b|\\bDx\\b|\\bDiagnosis\\b|\\bdx\\b|\\diagnosis", "", narrative,ignore.case = TRUE)) %>%
    unnest_tokens(word, narrative, token = "ngrams", n = 2)%>% #split each word as a row
  anti_join(stop_words)%>% #remove stop words

  count(word, sort = TRUE)

# remove stop words
pdf6 <- pdf5 %>%
  separate(word, into = c("first","second"), sep = " ", remove = FALSE) %>%
  anti_join(stop_words, by = c("first" = "word")) %>%
  anti_join(stop_words, by = c("second" = "word")) %>%
  filter(str_detect(first, "^[a-zA-Z]{3,}$") &
          str_detect(second, "^[a-zA-Z]{3,}$"))%>%
  mutate(percentage = n / sum(n) * 100) 
```
:::

> "Head Injury" is the most re-occurring pair of words in the Narrative data

```{r}
as.datatable(formattable(pdf6|>
                           filter(percentage>0.25),digits=2, list(
  #n = percent,
  n = color_tile("transparent", "#a1caf1")

)), rownames = F,
filter = 'top', 
options = list(
  pageLength = 10, autoWidth = F#,
  #order = list(list(4, 'desc'))#asc
),
class = 'bootstrap'
)
```

## Text Network Analysis

Text network analysis can be used to represent the narratives as a network graph. The words are the nodes and their co-occurrences are the relations. With the narratives encoded as a network, advanced graph theory algorithms can be used to detect the most influential keywords, identify the main topics, the relations between them, and get insights into the structure of the discourse. By taking this approach, the focus is on the relations between the words, while retaining contextual information and the narrative. Unlike bag-of-words, LDA-based, or Word2Vec models which may lose information about the words sequence, text network can be built in a way that retains the narrative and, therefore, provides more accurate information about the text and its topical structure.

```{r }
#| layout-ncol: 2
#| fig-dpi: 800
#| column: page
pdf_net = pdf6|>
  #filter(n>1000)|>
  filter(percentage>0.25)|>
  select(first, second)|>
  rename(from = first, to = second)

network = graph_from_data_frame(d = pdf_net)

network2 = toVisNetworkData(network)
n = data.frame(network2$nodes, font.size = 30)
e = data.frame(network2$edges)
visNetwork(n,e)|> 
  visIgraphLayout(layout = "layout_with_kk",#layout_on_grid, layout.star, layout_on_sphere, layout_with_kk
                  physics = F)|>#layout_in_circle
    visNodes(size = 30) |>
  #visEdges(arrows = "from")|>
  visOptions(highlightNearest = list(enabled = T, hover = T, degree =1),
             nodesIdSelection = T)


## pagerank
pagerank = network%>%
  as_tbl_graph() %>%
  mutate(pagerank = centrality_pagerank())|>
  as_tibble()%>%
  arrange(desc(pagerank))

as.datatable(formattable(pagerank, digits=2,list(
  pagerank = percent,
  pagerank = color_tile("transparent", "#a1caf1")

)), rownames = F,
filter = 'top', 
options = list(
  pageLength = 10, autoWidth = F#,
  #order = list(list(4, 'desc'))#asc
),
class = 'bootstrap'
)
```

## [Text Summarization](https://www.emilhvitfeldt.com/post/2018-03-15-tidy-text-summarization/)

This section builds on the previous, by leveraging [TextRank](https://web.eecs.umich.edu/~mihalcea/papers/mihalcea.emnlp04.pdf), which is based on the [PageRank](https://en.wikipedia.org/wiki/PageRank) algorithm to extract sentences i.e. extractive text summarization. in this analysis, sentences are modelled as the vertices and words as the connection edges. So sentences with words that appear in many other sentences are seen as more important.

::: {.callout-note collapse="true"}
```{r}
#https://www.emilhvitfeldt.com/post/2018-03-15-tidy-text-summarization/

pdf5b = pdf_0 %>%
  filter(age>80)|>
  mutate(narrative = 
           gsub("\\bYOF\\b|\\bYOM\\b|\\bPT\\b|\\bDX\\b|\\bdx\\b|\\bDiagnosis\\b|\\DX\\b|\\DX|\\dx|\\yof|\\yom|\\yf|\\ym", "", narrative,ignore.case = TRUE))|>
  head(50)

```

```{r}
article_sentences <-pdf5b %>%
  unnest_tokens(sentence, narrative, token = "sentences") %>%
  mutate(sentence_id = row_number()) %>%
  select(sentence_id, sentence)

article_words <- article_sentences %>%
  unnest_tokens(word, sentence)%>%
  anti_join(stop_words, by = "word")

#Running TextRank
article_summary <- textrank_sentences(data = article_sentences, 
                                      terminology = article_words)
```
:::

-   An Overview of the top 5 sentences based on the first 50 narratives for Adults older than 80 years

```{r}

#extracting the top 3 
article_summary[["sentences"]] %>%
  arrange(desc(textrank)) %>% 
  slice(1:5) %>%
  pull(sentence)

```

**Key takeaway(s)**

-   Fall accidents tend to occur at the nursing home, for adults older than 85 years of age

# Modelling: Identifying themes based on narrative embedding

In this section, two clustering algorithms [K-means](https://uc-r.github.io/kmeans_clustering) and [DBSCAN](http://www.sthda.com/english/wiki/wiki.php?id_contents=7940) were experimented with to test the efficacy in identifying theme clusters.

K-means clustering is the most commonly used unsupervised machine learning algorithm for partitioning a given data set into a set of k groups (i.e. k clusters), where k represents the number of groups pre-specified by the analyst.The basic idea behind k-means clustering consists of defining clusters so that the total intra-cluster variation (known as total within-cluster variation) is minimized.

DBSCAN is a density-based clustering algorithm, which can be used to identify clusters of any shape in data set containing noise and outliers. The key idea is that for each point of a cluster, the neighborhood of a given radius has to contain at least a minimum number of points.The goal is to identify dense regions, which can be measured by the number of objects close to a given point.

**Processing step**: Applying UMAP reduction step to the PCA processed data, to present data in 2-dimensional space.

::: panel-tabset
## K-Means

::: {.callout-note collapse="true"}
## Workflow

```{r}
set.seed(123)
#create recipe
recipe_object_2 = recipe(~.,-cpsc_case_number , data = emb2d)|>#emb2c
  step_umap(all_numeric_predictors(),-cpsc_case_number
            ) 
 
#extract table
kmeans_tbl = recipe_object_2|>
  prep()|>
  juice()


```

```{r}
k_tbl2 = kmeans_tbl|>
  slice_sample(n=1000)

#silhouette,mb 
fviz_nbclust(k_tbl2,
             FUNcluster = kmeans,
             method = c("silhouette"),
             diss = NULL,
             k.max = 10,
             nboot = 100)

#gap stats
fviz_nbclust(k_tbl2,
             FUNcluster = kmeans,
             method = c("gap_stat"),
             diss = NULL,
             k.max = 10,
             nboot = 10)

#elbow method
fviz_nbclust(k_tbl2,
             FUNcluster = kmeans,
             method = c("wss"),
             diss = NULL,
             k.max = 10,
             nboot = 10)
```

```{r}
#Specifying clustering models, arbitrarily set the number of clusters to 4
kmeans_spec_best_emb <- k_means(num_clusters = 4) %>% 
  set_engine("ClusterR")

#create workflow
kmeans_wf_best_emb <- workflow()|>
  add_recipe(recipe_object_2)|>
  add_model(kmeans_spec_best_emb)

#fit model
kmeans_best_fit_mdl_emb <- kmeans_wf_best_emb|>
  fit(data = emb2d) #emb2d

kmeans_best_fit_mdl_emb
```

```{r}
#predict cluster
pdf_cluster_emb =kmeans_best_fit_mdl_emb %>% #kmeans_fit_emb|>
  augment(emb2d) #emb2c

pdf_pca_emb = recipe_object_2 |>
  prep()|>
  juice()

#merge dataframe
pdf_cluster_emb_merge = pdf_cluster_emb|>
  select(cpsc_case_number, .pred_cluster)|>
  left_join( pdf_0, by = "cpsc_case_number")|>
  bind_cols(pdf_pca_emb)
```
:::

### Visualizing Clusters

```{r}
emb_plot = pdf_cluster_emb_merge %>%
 #filter(!(other_race %in% "")) %>%
 ggplot() +
 aes(x = UMAP1, y = UMAP2, colour = .pred_cluster, 
     #size = age, 
     text = activity) +
 geom_point(shape = "circle") +
 scale_color_hue(direction = 1) +
 #geom_mark_ellipse(aes(color = .pred_cluster), expand = unit(0.5,"mm"))+
  theme_minimal()
#emb_plot
ggplotly(emb_plot,tooltip = "text")
```

## DBSCAN

::: {.callout-note collapse="true"}
## Workflow

Identifying the optimal "eps" parameter

```{r}
#https://stats.stackexchange.com/questions/88872/a-routine-to-choose-eps-and-minpts-for-dbscan
pdf_pca_emb_db = pdf_pca_emb|> select(-cpsc_case_number)
dbscan::kNNdistplot(pdf_pca_emb_db, k =  450)
abline(h = 0.4, lty = 2)
```

```{r}
set.seed(123)
pdf_dbscan = dbscan(pdf_pca_emb_db, eps = 0.4, minPts = 550)
```

```{r}
# Plot DBSCAN results
#hullplot(pdf_pca_emb_db, pdf_dbscan$cluster)
```
:::

### Visualize Cluster

```{r}
pdf_cluster_emb_merge_db = pdf_cluster_emb_merge|>
  cbind(pdf_dbscan$cluster)|>
  rename( db_cluster = "pdf_dbscan$cluster")|>
  mutate(db_cluster = as_factor(db_cluster))

emb_plot_db = pdf_cluster_emb_merge_db %>%
 #filter(!(other_race %in% "")) %>%
 ggplot() +
 aes(x = UMAP1, y = UMAP2, colour = db_cluster, 
     #size = age, 
     text = activity) +
 geom_point(shape = "circle",size = 0.5) +
 scale_color_hue(direction = 1) +
 #geom_mark_ellipse(aes(color = .pred_cluster), expand = unit(0.5,"mm"))+
  theme_minimal()
#emb_plot
ggplotly(emb_plot_db,tooltip = "text")
```
:::

**Key takeaway**:

-   Effectiveness of Embeddings: Using embeddings proved effective in extracting themes.
-   DBSCAN vs. K-means: Density-based clustering (DBSCAN) demonstrated greater efficacy in identifying clusters compared to the k-means approach.
-   Theme Identification: Density-based clustering revealed the presence of 9 major themes, with cluster 0 being categorized as outlier/general themes

# Understanding themes

In this section, theme clusters from the density-based algorithm are explored[^3] in relation to the "activities" associated with falls.

[^3]: Note: Click on the legend to isolate a cluster theme or themes from others

```{r message=FALSE, warning=FALSE}
#| height: 5

pdf_cluster_emb_rdr_1bb <- pdf_cluster_emb_merge_db %>%
  #filter(.pred_cluster == "Cluster_5") %>%
  #group_by(.pred_cluster)|>
  group_by(db_cluster)|>
  count(activity, sort = TRUE) %>%
  mutate(percentage = n / sum(n) ) %>%
  filter(percentage > 0.02) %>%
  select(activity, percentage) |>
  pivot_wider(names_from = activity, values_from = percentage)|>
  mutate_all(~replace_na(.x, 0))

pdf_cluster_emb_rdr_1c = pdf_cluster_emb_rdr_1bb|>
  as_tibble()|>
  #mutate_each(funs(rescale), -.pred_cluster)
  mutate_each(funs(rescale), -db_cluster)



rd = pdf_cluster_emb_rdr_1c|>
  ggradar(
    #group.colours = palette_light()|>unname(),
    #fill = T,
    #fill.alpha = 0.1,
    plot.title = "Radar Chart",
    group.line.width = 1,
    group.point.size = 1,
    font.radar = "ariel" ,
    axis.label.size = 3,
    grid.label.size = 5
  )+ theme(
    text = element_text(family = "ariel"),
    plot.title = element_text(size = 12)
  ) #+ facet_wrap(~.pred_cluster,ncol = 3)
ggplotly(rd)
```

## Exploring clusters

-   Visualizing Insights on cluster 6

::: {.callout-note collapse="true"}
## Workflow

```{r}
selection = "6"
pdf_cluster_emb_1 = pdf_cluster_emb_merge_db %>% #pdf_cluster_emb_merge
  #filter(.pred_cluster == "Cluster_1")|>
  filter(db_cluster == selection)|>
  mutate(narrative = 
           gsub("\\byof\\b|\\byom\\b|\\bPT\\b|\\bdx\\b|\\bDiagnosis\\b|\\bhead\\b|\\bhip\\b|\\bleg\\b|\\bscalp\\b|\\bskin\\b|\\barm\\b|\\bknee\\b|\\belbow\\b|\\bshoulder\\b|\\bneck\\b|\\bchest\\b|\\bforehead\\b|\\bwrist\\b|\\brib\\b|\\bhit\\b|\\bhitting\\b|\\bclosed\\b", "", narrative,ignore.case = TRUE)) %>%
    unnest_tokens(word, narrative, token = "ngrams", n = 2)%>% #split each word as a row
  anti_join(stop_words)%>% #remove stop words

  count(word, sort = TRUE)%>%
  separate(word, into = c("first","second"), sep = " ", remove = FALSE) %>%
  anti_join(stop_words, by = c("first" = "word")) %>%
  anti_join(stop_words, by = c("second" = "word")) %>%
  filter(str_detect(first, "^[a-zA-Z]{3,}$") &
          str_detect(second, "^[a-zA-Z]{3,}$"))%>%
  mutate(percentage = n / sum(n) * 100)
```
:::

```{r}
pdf_cluster_emb_net_1 = pdf_cluster_emb_1|>
  #filter(n>20)|>
  filter(percentage>0.25)|>
  select(first, second,n)|>
  rename(from = first, to = second)

network_emb_c1 = graph_from_data_frame(d = pdf_cluster_emb_net_1)
network_emb_c1_2 = toVisNetworkData(network_emb_c1)
n = data.frame(network_emb_c1_2$nodes, font.size = 30)


e = data.frame(network_emb_c1_2$edges)
visNetwork(n,e)|> #, height = "600px",width = "600px"
  visIgraphLayout(layout = "layout_with_kk",#"layout.star",#"layout_with_kk",
                  physics = F)|>#layout_in_circle
    visNodes(size = 30) |>
  #visEdges(arrows = "from")|>
  visOptions(highlightNearest = list(enabled = T, hover = T, degree =1),
             nodesIdSelection = T)
```

```{r}
## pagerank
pagerank_1 = network_emb_c1%>%
  as_tbl_graph() %>%
  mutate(pagerank = centrality_pagerank())|>
  as_tibble()%>%
  arrange(desc(pagerank))

as.datatable(formattable(pagerank_1, digits=2,list(
  pagerank = percent,
  pagerank = color_tile("transparent", "#a1caf1")

)), rownames = F,
filter = 'top', 
options = list(
  pageLength = 10, autoWidth = F#,
  #order = list(list(4, 'desc'))#asc
),
class = 'bootstrap'
)
```

## Cluster Text Summarization

::: {.callout-note collapse="true"}
## Workflow

```{r}
pdf_sum_1 = pdf_cluster_emb_merge_db %>%
  #filter(.pred_cluster == "Cluster_1")|>
  filter(db_cluster == selection)|>
  mutate(narrative = 
           gsub("\\byof\\b|\\byom\\b|\\bPT\\b|\\bdx\\b|\\bDiagnosis\\b|\\dx\\b|\\dx|\\yof|\\yom|\\yf|\\ym|\\:", "", narrative,ignore.case = TRUE))|>
 slice_sample(n=100)#, by=c(sex,product_1,diagnosis)
```

```{r}
article_sentences_1 <-pdf_sum_1 %>%
  unnest_tokens(sentence, narrative, token = "sentences") %>%
  mutate(sentence_id = row_number()) %>%
  select(sentence_id, sentence)

article_words_1 <- article_sentences_1 %>%
  unnest_tokens(word, sentence)%>%
  anti_join(stop_words, by = "word")

#Running TextRank
article_summary_1 <- textrank_sentences(data = article_sentences_1, 
                                      terminology = article_words_1)
```
:::

```{r}
#extracting the top 3 
article_summary_1[["sentences"]] %>%
  arrange(desc(textrank)) %>% 
  slice(1:10) %>%
  pull(sentence)
```

> **Summary overview of all cluster themes**

| Cluster | Theme                                   | Associated Activities     | Obstacle                                 | Injury                                 | Top 3 Keywords (excluding the term "Fall") |
|------------|------------|------------|------------|------------|------------|
| 0       | General Elderly Falls and Injuries      | Lost Balance              | Ladders, others not specified            | Others                                 | left, admit, contusion                     |
| 1       | Head Injuries from Falls                | Standing, Rising          | Bed or bed-frames                        | Laceration                             | injury, laceration, contusion              |
| 2       | Falls Resulting in Shoulder Injuries    | Tripped, Exercise, Sports | Exercise                                 | Dislocation, Avulsion, Strain & Sprain | fracture, left, humerus                    |
| 3       | Hip Injuries from Falls                 | Rising, Tripped           | Footwear                                 | Fracture, Strain & Sprain              | fracture, left, femur                      |
| 4       | Syncope-Related Head Injuries           | Fainted                   | Toilets                                  | Laceration                             | syncope, laceration, striking              |
| 5       | Rib Injuries from Falls                 | Standing                  | Bath-tubs or Showers                     | Fracture                               | left, fracture, ribs                       |
| 6       | Alcohol-Related Head Injuries and Falls | Stair Navigation, others  | Stairs or steps                          | Poisoning, Laceration                  | alcohol, blood, intoxication               |
| 7       | Buttocks Contusions from Falls          | Rising, sitting, slipped  | bed or bed-frames                        | Contusions                             | contusions, buttocks, lower                |
| 8       | Atrial fibrillation related falls       | Sitting, Standing         | Tables, rugs & carpets, Ceilings & Walls | Hermatomia                             | encounter, laceration, initial             |
| 9       | Floor Falls and Associated Injuries     | Walking, Slipped          | Floors , balconies                       | Contusions                             | Falling, Floor, Dizzy                      |

## Further Exploration

In this section, building on the understanding of the cluster themes, these themes are further explored in relation to other variables like Age, severity[^4] level, sex etc

[^4]: see appendix for severity level definitions

::: panel-tabset
## Exploration 1: Age Distribution

```{r}
#Cluster themes by age distribution
plot_2 = ggplot(pdf_cluster_emb_merge_db) +
 aes(x = db_cluster, y = age) +
 geom_boxplot(fill = "#AEC8DF") +
 labs(x = "Cluster themes",title = "Cluster themes by age distribution") +
 theme_minimal()
ggplotly(plot_2)
```

## Exploration 2: Severity Level

```{r}
#Cluster themes by Severity levels
plot_3 =ggplot(pdf_cluster_emb_merge_db) +
 aes(x = db_cluster, fill = severity_level) +
 geom_bar() +
 scale_fill_brewer(palette = "Blues", 
 direction = 1) +
 labs(x = "Cluster themes", y = "Number of narrative", title = "Cluster themes by Severity levels", 
 caption = "..", fill = "Severity Level") +
 theme_minimal()
ggplotly(plot_3)
```

## Exploration 3: Sex

```{r}


plot_4 = ggplot(pdf_cluster_emb_merge_db) +
 aes(x = db_cluster, fill = sex) +
 geom_bar() +
 scale_fill_brewer(palette = "Blues", 
 direction = 1) +
 labs(x = "Cluster Themes", y = "Number of Narratives", title = "Cluster themes by sex") +
 theme_minimal()

ggplotly(plot_4)

```

## Exploration 4: Location

```{r}


plot_5 = pdf_cluster_emb_merge_db %>%
 filter(!(location %in% "UNK")) %>%
 ggplot() +
 aes(x = db_cluster, fill = location) +
 geom_bar() +
 scale_fill_brewer(palette = "Blues", 
 direction = 1) +
 labs(x = "Cluster Theme", y = "Number of Narratives", title = "Cluster themes by incident location") +
 theme_minimal()

ggplotly(plot_5)
```

## Exploration 5: Trends

```{r message=FALSE}
plot_6 = pdf_cluster_emb_merge_db|>
  group_by(treatment_date,db_cluster)|>
  summarise(
    cases = n()
  )


plot_6a = ggplot(plot_6) +
 aes(x = treatment_date, y = cases, colour = db_cluster) +
 geom_line() +
 scale_color_hue(direction = 1) +
 labs(y = "Cases", title = "Trend of Cluster Themes", color = "Cluster themes") +
 theme_minimal()


 ggplotly(plot_6a)
```

> Summary table view of the average number of cases for each cluster theme across the different years

```{r message=FALSE}
plot_7 = plot_6|> mutate(year = year(treatment_date))|>
  group_by(year,db_cluster)|>
  summarise(
    avg_cases = mean(cases)
  )|>
  pivot_wider(names_from = year, values_from = avg_cases)
plot_7$change_from_2021 =( plot_7$"2022"/plot_7$"2021")-1

as.datatable(formattable(plot_7, digits=2,list(
  change_from_2021 = percent,
  "2019" = color_tile("transparent", "#a1caf1"),
  "2020" = color_tile("transparent", "#a1caf1"),
  "2021" = color_tile("transparent", "#a1caf1"),
  "2022" = color_tile("transparent", "#a1caf1"),
  change_from_2021 = color_tile("transparent", "#a1caf1")

)), rownames = F,
filter = 'top', 
options = list(
  pageLength = 10, autoWidth = F#,
  #order = list(list(4, 'desc'))#asc
),
class = 'bootstrap'
)
```
:::

# Conclusion

-   Combining embeddings with dimensionality reduction techniques has proven to be highly effective in the extraction of cluster themes.
-   DBSCAN outperforms k-means in cluster identification.
-   Patients in the "*Alcohol-Related Head Injuries and Falls*" group tend to be younger, while the "*Atrial fibrillation related falls*" group was generally older
-   The "*Syncope-Related Head* *Injuries*" group had a higher rate of severe cases compared to other groups.
-   In comparison to previous year (2021), cases involving "*Head Injuries from Falls*", "*Syncope-Related Head* *Injuries*" and "*Rib Injuries from Falls*" saw the most significant increase in the average number of cases.

# Appendix

-   Disposition classification from which the *severity levels* were derived from

| **Disposition Code** | **Category** |
|----------------------|--------------|
| 1                    | Not Severe   |
| 2                    | Not Severe   |
| 4                    | Severe       |
| 5                    | Severe       |
| 6                    | Not Severe   |
| 8                    | Severe       |
| 9                    | Not Severe   |

In this classification:

-   "Severe" includes disposition codes 4 (Treated and admitted for hospitalization), 5 (Held for observation), and 8 (Fatality, including DOA and deaths in the ED or after admission).

-   "Not Severe" includes disposition codes 1 (Treated and released, or examined and released without treatment, or transfers for treatment to another department of the same facility without admission), 2 (Treated and transferred to another hospital), 6 (Left without being seen, Left against medical advice, Left without treatment, Eloped), and 9 (Not recorded).

```{r}
sessionInfo()
```
